% Chapter 2

% Main chapter title
\chapter{Traffic Prediction: Literature Review}

% For referencing the chapter elsewhere, use \ref{Chapter2}
\label{Chapter2}

% This is for the header on each page - perhaps a shortened title
\lhead{Chapter 2. \emph{Traffic Prediction: Literature Review}}

% Quotation
{``There is no way that we can predict the weather six months ahead beyond giving the seasonal
average"}
\begin{flushright}
Stephen Hawking, \textit{Black Holes and Baby Universes} (1993)
\end{flushright}

%---------------------------------------------------------------------------------------------------
%	CONTENT
%---------------------------------------------------------------------------------------------------
\section{Introduction}
%\cite{ahmed1979analysis,van2012short,vlahogianni2004short,vlahogianni2014short,smith1997traffic}
In this chapter we provide a reasonably complete review of existing literature on short term
traffic flow prediction. Research on short term traffic prediction has been going on since 1979,
\citet{ahmed1979analysis}. After more than three decades of research, short term traffic flow
prediction is still an interesting reserch subject for many professionals around the world. The
simplest reseaon being the complex non-linear nature of traffic data and the effects of
non-recurrent events(weather, public events, accidents etc.) on it.  Critical reviews of existing
literature on short term traffic flow have been presented in detail by \citet{smith1997traffic},
\citet{vlahogianni2004short}, \citet{van2012short} and \citet{vlahogianni2014short}. The use of
the phrase 'short term' limits the scope of traffic prediction in terms of the prediction horizon
which usually varies between few seconds to few hours depending upon the approach and application.

The traffic parameter that are predicted can be - flow(number of vehicles per hour), time(minutes
to travel between two points), speed(mean speed in km/hour) and density(number of vehicles per km).

Selecting the right model for short term traffic prediction is a challenging task. A number of
models have been suggested and yet there is no one-fits-all model. The various methods that have
been suggested for short term traffic prediction can be categorised into four groups -
na誰ve, parametric, non-parametric and hybrid as shown in figure \ref{fig:taxonomyTrafficPrediction}.


\begin{figure}
\centering
\begin{tikzpicture}
  \begin{scope}[blend group = soft light]
    \fill[red!30!white]   ( 90:1.2) circle (2);
    \fill[green!30!white] (210:1.2) circle (2);
    \fill[blue!30!white]  (330:1.2) circle (2);
  \end{scope}
  \node at (90:2)    {Na誰ve};
  \node at (210:2)   {Parametric};
  \node at (330:2)   {Non-Parametric};
  \node              {Hybrid};
\end{tikzpicture}
\caption{Methods used for short term traffic prediction} \label{fig:taxonomyTrafficPrediction}
\end{figure}

\section{Na誰ve methods}
These are heuristics methods, and often used in practice because of their simplicity and the ease
of implementations. In most cases these methods are used as baselines for comparison while
creating more advances methods. The simplest naive approach in short term prediction would be to
take the last observed value and this involves no computational effort. Another simple heuristic
method known as the historical averages uses the average of past observed values.

\section{Parametric models}
In parametric models, we estimate the parameters from the training dataset to determine the
function that classifies new unseen data. The number of parameters are fixed. The advantage of
parametric models are that these perform quite well in situations where the large amount of data
is not available. Some of the typical examples of parametric models include Linear and
nonlinear regression, ARIMA models, Kalman filter, Linear SVM etc.

\subsection{Classical regression}
In machine learning and statistical applications, the use of linear models are predominant. These
models are also important in time series domains such as traffic flow prediction. The primary
idea behind the regression is to express the output variable as a linear combination of input
vectors. We can express the linear regression in time series as an ouput influenced by a
collection of inputs, where the inputs could possibly be an independent series

        \begin{equation}
            x_{t} = \beta_{1}z_{t1} + \beta_{2}z_{t2} + ... + \beta_{q}z_{tq} + w_{t}
        \end{equation}

where $ \beta_{1}, \beta_{2},...,\beta_{q} $ are unknown regression coeffiecients and $w_{t}$ is
a random error.

\citet{hogberg1976estimation} used non-liner regression for traffic prediction.

\subsection{ARIMA}
%  Introduction to ARIMA models

ARIMA(Auto Regressive Integrated Moving Average) is a class of parametric regression models. In
this section we will introduce ARIMA and related methods such as exponential smoothing and moving
averages. For an in depth understanding of these models the reader is encouraged to refer to to
~\citet{tong1990non}, ~\citet{brockwell2006introduction} and ~\citet{box2015time}. It is
important to understand that ARIMA modelling works only with stationary time series data. A
stationary time series is one whose properties do not depend on the time it is being observed.
Trends and seasonality affect time series and hence make it non-stationary. Although this seems as a
big restriction, in short term traffic prediction, ARIMA models have been very successful. Two
basic models constituate ARIMA models - AR(autoregressive) and MA(moving average).

The main idea behind autoregressive models is that past values affect the present value, i.e.
$x_{t}$ can be expressed as a function of past p values $ x_{t-1}, x_{t-2},...,x_{t-p} $ , where
p is the number of steps into the past. We can express an autoregressive model of order p as below

        \begin{equation} \label{eq:autoregressive}
          x_{t} = \phi_{1}x_{t-1} + \phi_{2}x_{t-2} + ... + \phi_{p}x_{t-p} + w_{t}
        \end{equation}

where $x_{t}$ is stationary and $\phi_{1}, \phi_{2},..., \phi_{p}$ are constant
parameters that are to be chosen. We have added the term $w_{t}$ as a Guassian white noise with
zero mean and variance $\sigma^{2}_{w}$.

In the MA model, the current value is dependent on the last q one-step forecast errors
$e_{t-1}, e_{t-2},...,e_{t-q}$ and the white noise $w_{t}$. The expression for moving average
is

        \begin{equation} \label{eq:movingaverage}
          x_{t} = -\theta_{1}e_{t-1} - \theta_{2}e_{t-2} - ... - \theta_{q}e_{t-q} + w_{t}
        \end{equation}

$\theta_{1}, \theta_{2},..., \theta_{q}$ are the parameters to be chosen.

Now proceeding to an ARMA(autoregressive moving average) model, we define an ARMA(p,q) model
where the present value $x_{t}$ is dependent on p past recent values and q past recent forecast
errors and a white noise $w_{t}$.

        \begin{equation} \label{eq:arma}
          x_{t} = \phi_{1}x_{t-1} + \phi_{2}x_{t-2} + ... + \phi_{p}x_{t-p} - \theta_{1}e_{t-1}
          - \theta_{2}e_{t-2} - ... - \theta_{q}e_{t-q} + w_{t}
        \end{equation}

When q is 0, the model becomes an autoregressive model of order p, AR(p) and when p is 0 the model
is a moving average of order q, MA(q). We can rewrite \ref{eq:arma} by using the backshift
operator $B^{\alpha}$, which is defined as $B^{\alpha}z_{t} = z_{t-\alpha}$,

        \begin{equation} \label{eq:armarewrite}
          \phi(B)x_{t} = \theta(B)e_{t}
        \end{equation}

where
        \begin{equation}
            \phi(z) = 1 - \phi_{1}z - ... - \phi_{p}z^{p}
        \end{equation}
        \begin{equation}
            \theta(z) = 1 - \theta_{1}z - ... - \theta_{q}z^{q}
        \end{equation}

In practice, most time series data are non-stationary and so several approaches, for instance
by differencing, are taken to make it stationary before applying the ARMA(p,q) model. By
combining differencing with autoregressive and moving average we obtain the ARIMA model defined
as below
        \begin{equation} \label{eq:arima}
          x'_{t} = \phi_{1}x'_{t-1} + \phi_{2}x'_{t-2} + ... + \phi_{p}x'_{t-p} -
          \theta_{1}e_{t-1} - \theta_{2}e_{t-2} - ... - \theta_{q}e_{t-q} + w_{t}
        \end{equation}

where $x'_{t}$ is the differenced series. Formally the model is denotes as ARIMA(p,d,q) where p
is the order of autoregressive part, d is the degree of differencing and q is the order of moving
average. This is also known as a non-seasonal ARIMA model.

The common method used to determine the parameters in an ARIMA(p,d,q) model is known as the
Box-Jenkins approach (\citet{box2015time}) which is three stage procedure. The three stages are
identification, estimation and diagnostic checking. At the identification stage, the values p, d
and q are determined by observing the autocorrelation and partial autocorrelation functions of
the time series and its differences. At the estimation stage, the maximum liklihood estimates are
determined for each model parameter. Finally in the dignostics stage, the residuals are analysed
and model comparisions are done. If the model fits well then the standardised residuals behave as
an i.i.d. with mean zero and variance one.

% Application in traffic forecast
\citet{ahmed1979analysis} used Box-Jenkins method for short-term traffic forecast. The input data
used was 166 sets of time series traffic data collected by freeway traffic surveillance systems in
three locations - Los Angeles, Minneapolis and Detroit. The authors concluded an ARIMA(0,1,3)
model, based on the autocorrealtion and partial autocorrelation functions, as a resonable fit for
the short term prediction tasks for both traffic volume and occupancy. The model performance was
evaluated against a moving average, a double smoothing average and a Trigg and Leach adaptive
model. The comparisons suggest that the ARIMA model had better accuracy than the others. The
authors used this model in detecting traffic incidents by comparing the real-time flow occupancy
with the predicted value. \citet{nihan1980use} used the Box-Jenkins technique on monthly data
collected at 15 minutes interval on a freeway segment from 1968 to 1976 to forecast for the year
1977. After examining several models they finally selecte an ARIMA(12,1,7) model. The forecast
was done for average weekday volume with positive results.

\citet{williams2001multivariate} used an ARIMAX model to use upstream traffic data along with the
predicting location's traffic data while estimating the paramters of the ARIMA model. This is
done using ARIMAX model which is an extension of the ARIMA model where an exogenous variable is
used. The data was collected form four locations near Beaune, France. The data from three upstream
locations were used for forecasting at the fourth location in Beaune. The same data were used
in the proposed ATHENA and KARIMA models. The model was compared  against the univariate ARIMA,
ATHENA\citet{danech1991athena} and KARIMA(\citet{van1996combining}) models. The results show tha
the ARIMAX model consistently outperformed the ARIMA model. However the complexity of the ARIMAX
model is more than the ARIMA model with as many as twice the parameters to estimate. Also in case
of missing values the ARIMAX model performance degraded more than the ARIMA model.

\citet{min2009short} proposed a dynamic Space Time ARIMA (STARIMA) model for short term traffic
prediction. Their argument for the new proposed model was based on the factor that most of the
existed model failed to take the spatial information of the transporatation system into account.
The proposed dynamic STARIMA model combines STARIMA and Dynamic Turn Ratio Prediction (DTRP)
model. Using DTRP they dynamically updated the static matric $W_{k}$ in STARIMA model that contains
the structural information of the transporatation network. The results of the study showed
significant improvement in forecast accuracy. The authors later published another similar work
(\citet{min2010urban}) that used the generalised STARIMA (GSTARIMA) model.  The authors
presented the results where this model has a small improvements over the STARIMA model. However
the major drawbacks of the GSTARIMA model is the estimation of large number of parameters which
significantly increases the computational time. It also suffers in performance if enough historical
data is not available.

\citet{williams2003modeling} poposed for the acceptance of seasonal ARIMA models for short term
traffic prediction. A seasonal ARIMA $(p,d,q) (P,D,Q)_{s}$ for a time series {$x_{t}$} is one
where s is the period, d and D are nonnegative integers. The time series theorem known as the World
decomposition is used as the theoritical justification of applying seasonal ARIMA model to
univariate time series with stationarity. Data from two freeway locations, one each from the
United States and the United Kingdom were used for evaluating the model. The performance of the
models were compared against three heuristics approaches - historical averages, random walk and
deviation from historical avarages. The results show that for both the locations the seasonal
ARIMA has better performance than the three methods mentioned earlier. However the authors did
not present whether a non-seasonal ARIMA model would have similar performance. The only other
model that was considered for comparison was the KARIMA model, which did not perform as good as
the seasonal ARIMA model. \citet{kumar2015short} also used a seasonal ARIMA in the context of
limited data for short term traffic prediction. They used data collected over three days from an
arterial road in Chennai, India for the study. The model was validated on 24 hours ahead forecast
. Thier resluts were positive when compared with historical avereges and na誰ve methods. They
argued when availability of large traffic dataset is a constraint seasonal ARIMA method is a
better choice. \citet{szeto2009multivariate} used a hybrid SARIMA model with cell transmission
model for multivariate traffic prediction. The authors reasoned the use of multivariate models
captuered the spatial characteristics of the transporatation network and hence are the natural
and better choice over an univariate model. The model was validated against data collected form
the city center in Dublin, Ireland. The results at two junctions were compared against real
observations and had MAPE of 4.45 and 10.6. The authors however did not provide comparison
against other univariate models or multivariate models which could present the model's relative
performance.

% Conclusions - pros and cons
The major defficiency of tha ARIMA models is that they do not take the extremes into
consideration and focus on the means. This is in contrast to the nature of the traffic data.
ARIMA models are also have the inability to perform will with missing data as pointed out by
\citet{smith1997traffic}.


\subsection{Kalman filter}
Kalman filter is a paramteric regression technique usually used in the field of automatic control
systems and signal preocessing. It was proposed by \citet{kalman1960new}. It can be used on both
stationary and non-stationary time series.

\citet{okutani1984dynamic} used Kalman filtering in traffic prediction in an urabn network and
extended it for freeways.

\citet{wang2005real}

\citet{xie2007short}

\citet{guo2010real}

\citet{guo2014adaptive}

The main advantage of Kalman filtering is the state variable is updated continuously.

\subsection{Exponential smoothing}
In exponential smoothing method the forecast is the weighted average of past observations, while
the weights decrease exponentially for older observations. For time series data with no trends
and seasons single exponential smoothing is usually used.


\section{Non-Parametric models}
In nonparamtric models the parameters are not fixed, and vary with the amount of data available.
Usually more data is required for this models than parametric models. The advantage of these models
is that they can model the complex non-linear data better. Some of the widely used non-parametric
models are - k-Nearest Neighbour, Non-parametric regrssion and Neural Networks

\subsection{K-nearest neighbour}

The basic process of the k-nearest neighbour algorithm is described in figure \ref{fig:KnnProcessFlow}.


\tikzstyle{block} = [rectangle, draw, text width=5em, text centered, rounded corners,
minimum height=4em]
\tikzstyle{line} = [draw, -latex']
\tikzstyle{cloud} = [draw, ellipse, node distance=4cm, minimum height=3em]

\begin{figure}
\centering
\begin{tikzpicture}[node distance = 3cm, auto]
    % Place nodes
    \node [block] (pp) {Preprocess Data};
    \node [cloud, left of=pp] (hd) {Historical data};
    \node [cloud, right of=pp] (rd) {Real-time data};
    \node [block, below of=pp] (msv) {Match state vector};
    \node [block, below of=msv] (knn) {K nearest neighbour};
    \node [block, left of=knn, node distance=3cm] (prd) {Predictions};
    % Draw edges
    \path [line] (pp) -- (msv);
    \path [line] (msv) -- (knn);
    \path [line] (knn) -- (prd);
    \path [line,dashed] (hd) -- (pp);
    \path [line,dashed] (rd) -- (pp);
\end{tikzpicture}
\caption{K nearest neighbour process flow} \label{fig:KnnProcessFlow}
\end{figure}

\citet{lv2009real}

\citet{myung2011travel}

\citet{zhang2013improved}

\citet{meng2015two}


\subsection{Neural networks}
\label{subsec:neuralNetworksTrafficPred}
Artificial Neural Networks(ANN) were mathematical models (\citet{mcculloch1943logical},
\citet{rosenblatt1958perceptron}) designed to  provide a representation of how the human brain
works. It is obvious now that these mathematical models bear little resemblance to the structure
of brain, yet they have been hugely successful. Because they were initially inspired by the
biological brain, the term neural is associated with such kind of mathematical models. A basic
artificial neural network consists of a set of nodes connnected by edges with weights. We can say
that the nodes represent the biological neurons and the edges represent the synapses. The
conections among the nodes can be cyclic or acyclic. The former is known as a feedforward neural
network and the later as a recurrent network. We describe about these neural networks in more
details in chpater \ref{Chapter4}. Several variations of artificial neural networks have been
used in short term traffic prediction. Some well known examples include - \textit{Multilayer
perceptrons, Radial basis function networks, Kohnen maps} and \textit{Hopfield networks}.

\cite{dougherty1997short} apllied a backpropagation feedforward neural network in traffic flow
and speed predictions.

\subsection{Fuzzy logic}

\cite{zhang2008short}

\subsection{Bayesian networks}

\cite{castillo2008predicting}


\section{Hybrid Methods}
In recent years many hybrid methods have been tried in short term traffic prediction with mixed
results.

A hybrid method by combining kohonen maps with ARIMA model was proposed by \citet{van1996combining}.
The model known as KARIMA, used the same data(collected near Beaune, France) that was used in the
ATHENA model for an accurate comparison with the later.

\citet{chen2011short} proposed an ARIMA-GARCH model for short term traffic prediction. The
performance of this hybrid model when compared to the standard ARIMA model did not yield positive
results.

\section{Comparisons}
